---
title: "Final_Report.rmd"
author: "Kevin Roldan, Linda Zavala, Negeen Agharokh"
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(psych)
```


# Day 1: Project Setup & Data Import 

```{r}

library(readr)
NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv")
View(NHIS_Data_2021)
str(NHIS_Data_2021)
summary(NHIS_Data_2021)
head(NHIS_Data_2021)
```

# Day 2: Data Cleaning & Management

```{r}
##Day 2: Data Cleaning & Management
#Task 1: Variable Selection & Missing Value Check
library(tidyverse) #loaded tidyverse

NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv") #loaded dataset
summary(NHIS_Data_2021) #loaded summary of data to see range of variables

NHIS_vars <- NHIS_Data_2021 %>% #created subset of cleaned dataset with only needed variables
  select(AGEP_A, WEIGHTLBTC_A, HEIGHTTC_A, SEX_A, HISPALLP_A, EDUCP_A, PHSTAT_A, LSATIS4R_A) %>% #selected only variables needed
  filter( #used filter to remove missing values per codebook #! removes unwanted values
    AGEP_A >= 18 & AGEP_A <= 85 & !AGEP_A %in% c(97, 98, 99), #Age range from 18-85
    WEIGHTLBTC_A >= 100 & WEIGHTLBTC_A <= 299 & !WEIGHTLBTC_A %in% c(996, 997, 998, 999), #Weight range from 100-299
    HEIGHTTC_A >= 59 & HEIGHTTC_A <= 76 & !HEIGHTTC_A %in% c(96, 97, 98, 99), #Height range from 59-76
    SEX_A >= 1 & SEX_A <= 2 & !SEX_A %in% c(7, 9), #Sex range from 1-2
    HISPALLP_A >= 1 & HISPALLP_A <= 7 & !HISPALLP_A %in% c(97, 98, 99), #Race/ethnicity range from 1-7
    EDUCP_A >= 0 & EDUCP_A <= 10 & !EDUCP_A %in% c(97, 98, 99), #Education range from 0-10
    PHSTAT_A >= 1 & PHSTAT_A <= 5 & !PHSTAT_A %in% c(7, 8, 9), #General health status range from 1-5
    LSATIS4R_A >= 1 & LSATIS4R_A <= 4 & !LSATIS4R_A %in% c(7, 8, 9)) #Quality of life range from 1-4

summary(NHIS_vars) #loaded summary of cleaned data to verify all variables are within codebook range & missing values removed

#Task 2: Recoding Variables
NHIS_clean <- NHIS_vars %>% #created final cleaned dataset and added recoded education variable
  mutate(
    EDUCP_recode = case_when( #recoded education variable based on NHIS codebook categories
      EDUCP_A >= 0 & EDUCP_A <= 3 ~ 1, #0-3 Less Than High School
      EDUCP_A == 4 ~ 2, #4 High School Graduate
      EDUCP_A >= 5 & EDUCP_A <= 7 ~ 3, #5-7 Some College Education
      EDUCP_A >= 8 & EDUCP_A <= 10 ~ 4 #8-10 College Graduate or Better
    )
  )

NHIS_clean <- NHIS_clean %>%
  mutate(
    EDUCP_recode = factor( #converted recoded education variable into a labeled factor
      EDUCP_recode,
      levels = c(1, 2, 3, 4), #defined category order
      labels = c( #labeled each category in order
        "Less than High School",
        "High School Graduate",
        "Some College Education",
        "College Graduate or Better"
      )
    )
  )

summary(NHIS_clean) #loaded summary of cleaned dataset to confirm recoding worked.

write.csv(NHIS_clean, "Data/NHIS_clean.csv", row.names = FALSE) #saved final cleaned dataset as CSV file

```

# Day 3: Univariate & Bivariate Analysis & Visualization

```{r}
# 03_univariate_analysis.R

library(tidyverse)
NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv")

summary(NHIS_Data_2021$AGEP_A)
summary(NHIS_Data_2021$WEIGHTLBTC_A)
summary(NHIS_Data_2021$HEIGHTTC_A)

#Extra Stats
mean(NHIS_Data_2021$AGEP_A, na.rm = TRUE)
sd(NHIS_Data_2021$AGEP_A, na.rm = TRUE)

#Histograms
hist(NHIS_Data_2021$AGEP_A, main="Age Distribution", xlab="Age")
hist(NHIS_Data_2021$WEIGHTLBTC_A, main="Weight Distribution", xlab="Weight")
hist(NHIS_Data_2021$HEIGHTTC_A, main="Height Distribution", xlab="Height")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = AGEP_A)) +
  geom_histogram(bins = 30) +
  labs(title="Age Distribution", x="Age", y="Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = WEIGHTLBTC_A)) +
  geom_histogram(bins = 30) +
  labs(title="Weight Distribution", x="Weight", y="Count")
```
```{R}
ggplot(NHIS_Data_2021, aes(x = HEIGHTTC_A)) +
  geom_histogram(bins = 30) +
  labs(title="Height Distribution", x="Height", y="Count")
```
```{r}
#Boxplots
boxplot(NHIS_Data_2021$AGEP_A, main="Age Boxplot")
boxplot(NHIS_Data_2021$WEIGHTLBTC_A, main="Weight Boxplot")
boxplot(NHIS_Data_2021$HEIGHTTC_A, main="Height Boxplot")
```
```{r}
ggplot(NHIS_Data_2021, aes(y = AGEP_A)) +
  geom_boxplot() +
  labs(title="Age Boxplot")
```
```{r}
ggplot(NHIS_Data_2021, aes(y = WEIGHTLBTC_A)) +
  geom_boxplot() +
  labs(title="Weight Boxplot")
```
```{r}
ggplot(NHIS_Data_2021, aes(y = HEIGHTTC_A)) +
  geom_boxplot() +
  labs(title="Height Boxplot")
```
```{r}
#Qualitative Variables
#Frequency Tables
table(NHIS_Data_2021$SEX_A)
table(NHIS_Data_2021$HISPALLP_A)
table(NHIS_Data_2021$EDUCP_A)
table(NHIS_Data_2021$PHSTAT_A)
table(NHIS_Data_2021$LSATIS4R_A)
```
```{r}
#Bar Plots
barplot(table(NHIS_Data_2021$SEX_A), main="Sex Distribution")
barplot(table(NHIS_Data_2021$HISPALLP_A), main="Hispanic Status")
barplot(table(NHIS_Data_2021$EDUCP_A), main="Education Level")
barplot(table(NHIS_Data_2021$PHSTAT_A), main="Health Status")
barplot(table(NHIS_Data_2021$LSATIS4R_A), main="Life Satisfaction")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(SEX_A))) +
  geom_bar() +
  labs(title="Sex Distribution", x="Sex", y="Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(HISPALLP_A))) +
  geom_bar() +
  labs(title="Hispanic Status", x="Category", y="Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(EDUCP_A))) +
  geom_bar() +
  labs(title="Education Level", x="Category", y="Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(PHSTAT_A))) +
  geom_bar() +
  labs(title="Health Status", x="Category", y="Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(LSATIS4R_A))) +
  geom_bar() +
  labs(title="Life Satisfaction", x="Category", y="Count")
```
```{r}
#Qualitative Variables
#Frequency tables
table(NHIS_Data_2021$SEX_A)
table(NHIS_Data_2021$HISPALLP_A)
table(NHIS_Data_2021$EDUCP_A)
table(NHIS_Data_2021$PHSTAT_A)
table(NHIS_Data_2021$LSATIS4R_A)
```
```{r}
#Bar Plots
barplot(table(NHIS_Data_2021$SEX_A),
        main = "Sex Distribution",
        xlab = "Sex", ylab = "Count")
```
```{r}
barplot(table(NHIS_Data_2021$HISPALLP_A),
        main = "Hispanic Status",
        xlab = "Category", ylab = "Count")
```
```{r}
barplot(table(NHIS_Data_2021$EDUCP_A),
        main = "Education Level",
        xlab = "Category", ylab = "Count")
```
```{r}
barplot(table(NHIS_Data_2021$PHSTAT_A),
        main = "Health Status",
        xlab = "Category", ylab = "Count")
```
```{r}
barplot(table(NHIS_Data_2021$LSATIS4R_A),
        main = "Life Satisfaction",
        xlab = "Category", ylab = "Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(SEX_A))) +
  geom_bar() +
  labs(title = "Sex Distribution", x = "Sex", y = "Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(HISPALLP_A))) +
  geom_bar() +
  labs(title = "Hispanic Status", x = "Category", y = "Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(EDUCP_A))) +
  geom_bar() +
  labs(title = "Education Level", x = "Category", y = "Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(PHSTAT_A))) +
  geom_bar() +
  labs(title = "Health Status", x = "Category", y = "Count")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(LSATIS4R_A))) +
  geom_bar() +
  labs(title = "Life Satisfaction", x = "Category", y = "Count")
```
#04_bivariate_analysis.R

library(tidyverse)
NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv")
```{r}
#Quantitative vs. Qualitative:
boxplot(AGEP_A ~ SEX_A,
        data = NHIS_Data_2021,
        main = "Age by Sex",
        xlab = "Sex",
        ylab = "Age")
```
```{r}
boxplot(AGEP_A ~ EDUCP_A,
        data = NHIS_Data_2021,
        main = "Age by Education Level",
        xlab = "Education Level",
        ylab = "Age")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(SEX_A), y = AGEP_A)) +
  geom_boxplot() +
  labs(title = "Age by Sex",
       x = "Sex",
       y = "Age")
```
```{r}
ggplot(NHIS_Data_2021, aes(x = factor(EDUCP_A), y = AGEP_A)) +
  geom_boxplot() +
  labs(title = "Age by Education Level",
       x = "Education Level",
       y = "Age")
```
```{r}
#Qualitative vs. Qualitative:
ggplot(NHIS_Data_2021, 
       aes(x = factor(PHSTAT_A), fill = factor(LSATIS4R_A))) +
  geom_bar(position = "dodge") +
  labs(title = "General Health vs Life Satisfaction",
       x = "General Health",
       y = "Count",
       fill = "Life Satisfaction")
```
#Quantitative vs. Quantitative:
plot(NHIS_Data_2021$HEIGHTTC_A,
     NHIS_Data_2021$WEIGHTLBTC_A,
     main = "Height vs Weight",
     xlab = "Height",
     ylab = "Weight")
```{r}
ggplot(NHIS_Data_2021, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A)) +
  geom_point(alpha = 0.5) +
  labs(title = "Height vs Weight",
       x = "Height",
       y = "Weight")
```
cor(NHIS_Data_2021$HEIGHTTC_A,
    NHIS_Data_2021$WEIGHTLBTC_A,
    use = "complete.obs")

```

There is a strong positive relationship between height and weight, indicating that as height increases, weight also tends to increase. The high correlation coefficient (r = 0.77) suggests a strong linear association between the two variables.

# Day 4: Multivariate Analysis & Report Drafting

```{r}

multi_data <- NHIS_Data_2021 %>%
  select(HEIGHTTC_A, WEIGHTLBTC_A, SEX_A, EDUCP_A, PHSTAT_A, AGEP_A) %>%
  na.omit()

multi_data$SEX_A <- as.factor(multi_data$SEX_A)
multi_data$EDUCP_A <- as.factor(multi_data$EDUCP_A)
multi_data$PHSTAT_A <- as.factor(multi_data$PHSTAT_A)

ggplot(multi_data, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A, color = SEX_A)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ EDUCP_A) +
  labs(
    title = "Height vs Weight by Sex and Education",
    x = "Height (cm)",
    y = "Weight (lbs)",
    color = "Sex"
  ) +
  theme_minimal()

```


# Day 5: Finalizing, Communicating & Submission

```{r}


```




