---
title: "Final_Report.rmd"
author: "Kevin Roldan, Linda Zavala, Negeen Agharokh"
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(psych)
```




```{r}
# Day 1: Project Setup & Data Import 

library(readr) #loaded readr package for data import functions
NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv") #imported NHIS 2021 dataset from Data folder
View(NHIS_Data_2021) #opened dataset in spreadsheet view
str(NHIS_Data_2021) #displayed structure of dataset
summary(NHIS_Data_2021) #generated summary statistics for all variables to check ranges and missing values
head(NHIS_Data_2021) #displayed first 6 rows of dataset
```



```{r}
##Day 2: Data Cleaning & Management

#Task 1: Variable Selection & Missing Value Check
library(tidyverse) #loaded tidyverse

NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv") #loaded dataset
summary(NHIS_Data_2021) #loaded summary of data to see range of variables

NHIS_vars <- NHIS_Data_2021 %>% #created subset of cleaned dataset with only needed variables
  select(AGEP_A, WEIGHTLBTC_A, HEIGHTTC_A, SEX_A, HISPALLP_A, EDUCP_A, PHSTAT_A, LSATIS4R_A) %>% #selected only variables needed
  filter( #used filter to remove missing values per codebook #! removes unwanted values
    AGEP_A >= 18 & AGEP_A <= 85 & !AGEP_A %in% c(97, 98, 99), #Age range from 18-85
    WEIGHTLBTC_A >= 100 & WEIGHTLBTC_A <= 299 & !WEIGHTLBTC_A %in% c(996, 997, 998, 999), #Weight range from 100-299
    HEIGHTTC_A >= 59 & HEIGHTTC_A <= 76 & !HEIGHTTC_A %in% c(96, 97, 98, 99), #Height range from 59-76
    SEX_A >= 1 & SEX_A <= 2 & !SEX_A %in% c(7, 9), #Sex range from 1-2
    HISPALLP_A >= 1 & HISPALLP_A <= 7 & !HISPALLP_A %in% c(97, 98, 99), #Race/ethnicity range from 1-7
    EDUCP_A >= 0 & EDUCP_A <= 10 & !EDUCP_A %in% c(97, 98, 99), #Education range from 0-10
    PHSTAT_A >= 1 & PHSTAT_A <= 5 & !PHSTAT_A %in% c(7, 8, 9), #General health status range from 1-5
    LSATIS4R_A >= 1 & LSATIS4R_A <= 4 & !LSATIS4R_A %in% c(7, 8, 9)) #Quality of life range from 1-4

summary(NHIS_vars) #loaded summary of cleaned data to verify all variables are within codebook range & missing values removed

#Task 2: Recoding Variables
NHIS_clean <- NHIS_vars %>% #created final cleaned dataset and added recoded education variable
  mutate(
    EDUCP_recode = case_when( #recoded education variable based on NHIS codebook categories
      EDUCP_A >= 0 & EDUCP_A <= 3 ~ 1, #0-3 Less Than High School
      EDUCP_A == 4 ~ 2, #4 High School Graduate
      EDUCP_A >= 5 & EDUCP_A <= 7 ~ 3, #5-7 Some College Education
      EDUCP_A >= 8 & EDUCP_A <= 10 ~ 4 #8-10 College Graduate or Better
    )
  )

NHIS_clean <- NHIS_clean %>%
  mutate(
    EDUCP_recode = factor( #converted recoded education variable into a labeled factor
      EDUCP_recode,
      levels = c(1, 2, 3, 4), #defined category order
      labels = c( #labeled each category in order
        "Less than High School",
        "High School Graduate",
        "Some College Education",
        "College Graduate or Better"
      )
    )
  )

summary(NHIS_clean) #loaded summary of cleaned dataset to confirm recoding worked.

write.csv(NHIS_clean, "Data/NHIS_clean.csv", row.names = FALSE) #saved final cleaned dataset as CSV file

```



```{r}
##Day 3: Univariate & Bivariate Analysis & Visualization

library(tidyverse) #load tidyverse packages
NHIS_clean <- read.csv("Data/NHIS_clean.csv")  #import dataset

summary(NHIS_clean$AGEP_A) #summary stats on mean,median, mode...
summary(NHIS_clean$WEIGHTLBTC_A) #summary stats for weight
summary(NHIS_clean$HEIGHTTC_A)  #summary stats for height

#Extra Stats
mean(NHIS_clean$AGEP_A, na.rm = TRUE)  #average age while removing missing values
sd(NHIS_clean$AGEP_A, na.rm = TRUE)  #standard deviation

#Histograms
hist(NHIS_clean$AGEP_A, main="Age Distribution", xlab="Age") #histogram for age
hist(NHIS_clean$WEIGHTLBTC_A, main="Weight Distribution", xlab="Weight")
hist(NHIS_clean$HEIGHTTC_A, main="Height Distribution", xlab="Height")
```
The code loads the dataset and explores key variables (age, weight, and height) by generating summary statistics, calculating additional measures for age, and creating histograms to visualize their distributions.The histogram shows that most weight values cluster between about 120 and 250 pounds, indicating a right-skewed distribution, while a small number of extremely high values near 1000 pounds suggest strong outliers or possible data entry errors.
```{r}
ggplot(NHIS_clean, aes(x = AGEP_A)) + #plot using dataset
  geom_histogram(bins = 30) + #histogram within 30 bins
  labs(title="Age Distribution", x="Age", y="Count")  #title 
```
The code uses ggplot2 to create a histogram of the age variable, displaying its distribution with labeled axes and a descriptive title.The histogram shows that ages are fairly spread out across a wide range (roughly 20 to 80), indicating a relatively even distribution, with a slight concentration in middle-aged groups and a few extreme values near the upper end suggesting potential outliers.
```{r}
ggplot(NHIS_clean, aes(x = WEIGHTLBTC_A)) + #ggplot
  geom_histogram(bins = 30) +  #histogram of weight within 30 bins
  labs(title="Weight Distribution", x="Weight", y="Count") #add tittle
```
The code uses ggplot2 to visualize the distribution of weight by creating a histogram with appropriate labels and a descriptive title.The histogram shows that most weight values are concentrated between about 120 and 250 pounds, forming a right-skewed distribution, while extremely high values near 1000 pounds indicate clear outliers or likely data entry errors.
```{R}
ggplot(NHIS_clean, aes(x = HEIGHTTC_A)) + #ggplot
  geom_histogram(bins = 30) + #histogram of height within 30 bins
  labs(title="Height Distribution", x="Height", y="Count") #add tittle
```
The code uses ggplot2 to create a histogram that visualizes the distribution of height, with clear axis labels and a descriptive title.The histogram shows that most height values are concentrated between about 60 and 72 inches, indicating a roughly normal distribution, while a small group of unusually high values (around 95–100 inches) suggests the presence of potential outliers or data entry errors.
```{r}
#Boxplots
boxplot(NHIS_clean$AGEP_A, main="Age Boxplot") #boxplot for age
boxplot(NHIS_clean$WEIGHTLBTC_A, main="Weight Boxplot") #boxplot for weight
boxplot(NHIS_clean$HEIGHTTC_A, main="Height Boxplot") #boxplot for height
```
The code generates boxplots for age, weight, and height to visualize their distributions, highlighting the median, spread, and any potential outliers in each variable.
```{r}
ggplot(NHIS_clean, aes(y = AGEP_A)) + #ggplot on y axis
  geom_boxplot() + #boxplot for age including median, quartiles, outliers
  labs(title="Age Boxplot") #add tittle
```
The code uses ggplot2 to create a boxplot of age, allowing visualization of its distribution, central tendency, and potential outliers.The boxplot shows that the median age is around the mid-50s, with the middle 50% of values spread roughly between the late 30s and late 60s, indicating moderate variability, and no extreme outliers are clearly visible.
```{r}
ggplot(NHIS_clean, aes(y = WEIGHTLBTC_A)) + #ggplot on y axis
  geom_boxplot() + #box plot to distribute data
  labs(title="Weight Boxplot") #tittle
```
The code uses ggplot2 to create a boxplot of weight, visualizing its distribution, central tendency, and any potential outliers.The boxplot shows that most weight values are concentrated between roughly 140 and 220 pounds with a median around 180, indicating moderate spread, while a very extreme outlier near 1000 pounds highlights a likely data entry error or unusual value.
```{r}
ggplot(NHIS_clean, aes(y = HEIGHTTC_A)) + #ggplot to y axis
  geom_boxplot() + #distribution to height
  labs(title="Height Boxplot") #add tittle
```
The code uses ggplot2 to create a boxplot of height, allowing visualization of its distribution, central tendency, and potential outliers.The boxplot shows that most height values fall between about 64 and 70 inches with a median around 67 inches, indicating low to moderate variability, while several high values near 95–100 inches appear as outliers, suggesting possible data entry errors.
```{r}
#Qualitative Variables
#Kevin
#Frequency Tables
table(NHIS_clean$SEX_A)
table(NHIS_clean$HISPALLP_A)
table(NHIS_clean$EDUCP_recode)
table(NHIS_clean$PHSTAT_A)
table(NHIS_clean$LSATIS4R_A)
```
```{r}
#Bar Plots
barplot(table(NHIS_clean$SEX_A), main="Sex Distribution")
barplot(table(NHIS_clean$HISPALLP_A), main="Hispanic Status")
barplot(table(NHIS_clean$EDUCP_recode), main="Education Level")
barplot(table(NHIS_clean$PHSTAT_A), main="Health Status")
barplot(table(NHIS_clean$LSATIS4R_A), main="Life Satisfaction")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(SEX_A))) +
  geom_bar() +
  labs(title="Sex Distribution", x="Sex", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(HISPALLP_A))) +
  geom_bar() +
  labs(title="Hispanic Status", x="Category", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(EDUCP_recode))) +
  geom_bar() +
  labs(title="Education Level", x="Category", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(PHSTAT_A))) +
  geom_bar() +
  labs(title="Health Status", x="Category", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(LSATIS4R_A))) +
  geom_bar() +
  labs(title="Life Satisfaction", x="Category", y="Count")
```
```{r}
#Qualitative Variables
#Frequency tables
table(NHIS_clean$SEX_A)
table(NHIS_clean$HISPALLP_A)
table(NHIS_clean$EDUCP_recode)
table(NHIS_clean$PHSTAT_A)
table(NHIS_clean$LSATIS4R_A)
```
```{r}
#Bar Plots
barplot(table(NHIS_clean$SEX_A),
        main = "Sex Distribution",
        xlab = "Sex", ylab = "Count")
```
```{r}
barplot(table(NHIS_clean$HISPALLP_A),
        main = "Hispanic Status",
        xlab = "Category", ylab = "Count")
```
```{r}
barplot(table(NHIS_clean$EDUCP_recode),
        main = "Education Level",
        xlab = "Category", ylab = "Count")
```
```{r}
barplot(table(NHIS_clean$PHSTAT_A),
        main = "Health Status",
        xlab = "Category", ylab = "Count")
```
```{r}
barplot(table(NHIS_clean$LSATIS4R_A),
        main = "Life Satisfaction",
        xlab = "Category", ylab = "Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(SEX_A))) +
  geom_bar() +
  labs(title = "Sex Distribution", x = "Sex", y = "Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(HISPALLP_A))) +
  geom_bar() +
  labs(title = "Hispanic Status", x = "Category", y = "Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(EDUCP_recode))) +
  geom_bar() +
  labs(title = "Education Level", x = "Category", y = "Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(PHSTAT_A))) +
  geom_bar() +
  labs(title = "Health Status", x = "Category", y = "Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(LSATIS4R_A))) +
  geom_bar() +
  labs(title = "Life Satisfaction", x = "Category", y = "Count")
```
```{r}
#04_bivariate_analysis.R 
#Linda

library(tidyverse)
NHIS_clean <- read.csv("Data/NHIS_clean.csv")

#Quantitative vs. Qualitative:
boxplot(AGEP_A ~ SEX_A,
        data = NHIS_clean,
        main = "Age by Sex",
        xlab = "Sex",
        ylab = "Age")
```
```{r}
boxplot(AGEP_A ~ EDUCP_recode,
        data = NHIS_clean,
        main = "Age by Education Level",
        xlab = "Education Level",
        ylab = "Age")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(SEX_A), y = AGEP_A)) +
  geom_boxplot() +
  labs(title = "Age by Sex",
       x = "Sex",
       y = "Age")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(EDUCP_recode), y = AGEP_A)) +
  geom_boxplot() +
  labs(title = "Age by Education Level",
       x = "Education Level",
       y = "Age")
```
```{r}
#Qualitative vs. Qualitative:
ggplot(NHIS_clean, 
       aes(x = factor(PHSTAT_A), fill = factor(LSATIS4R_A))) +
  geom_bar(position = "dodge") +
  labs(title = "General Health vs Life Satisfaction",
       x = "General Health",
       y = "Count",
       fill = "Life Satisfaction")
```
#Quantitative vs. Quantitative:
plot(NHIS_clean$HEIGHTTC_A,
     NHIS_clean$WEIGHTLBTC_A,
     main = "Height vs Weight",
     xlab = "Height",
     ylab = "Weight")
```{r}
ggplot(NHIS_clean, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A)) +
  geom_point(alpha = 0.5) +
  labs(title = "Height vs Weight",
       x = "Height",
       y = "Weight")
```
cor(NHIS_clean$HEIGHTTC_A,
    NHIS_clean$WEIGHTLBTC_A,
    use = "complete.obs")

#There is a strong positive relationship between height and weight, indicating that as height increases, weight also tends to increase. The high correlation coefficient (r = 0.77) suggests a strong linear association between the two variables.
```



```{r}
##Day 4: Multivariate Analysis & Report Drafting

multi_data <- NHIS_clean %>%
  select(HEIGHTTC_A, WEIGHTLBTC_A, SEX_A, EDUCP_recode, PHSTAT_A, AGEP_A) %>%
  na.omit()

multi_data$SEX_A <- as.factor(multi_data$SEX_A)
multi_data$EDUCP_recode <- as.factor(multi_data$EDUCP_recode)
multi_data$PHSTAT_A <- as.factor(multi_data$PHSTAT_A)

ggplot(multi_data, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A, color = SEX_A)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ EDUCP_recode) +
  labs(
    title = "Height vs Weight by Sex and Education",
    x = "Height (cm)",
    y = "Weight (lbs)",
    color = "Sex"
  ) +
  theme_minimal()

```



```{r}
# Day 5: Finalizing, Communicating & Submission

```




