---
title: "Final_Report.rmd"
author: "Kevin Roldan, Linda Zavala, Negeen Agharokh"
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(psych)
```



# Day 1: Project Setup & Data Import 

```{r}
library(readr) #loaded readr package for data import functions
NHIS_Data_2021 <- read_csv("Data/NHIS_Data_2021.csv") #imported NHIS 2021 dataset from Data folder
View(NHIS_Data_2021) #opened dataset in spreadsheet view
str(NHIS_Data_2021) #displayed structure of dataset
summary(NHIS_Data_2021) #generated summary statistics for all variables to check ranges and missing values
head(NHIS_Data_2021) #displayed first 6 rows of dataset
```


# Day 2: Data Cleaning & Management
```{r}

#Task 1: Variable Selection & Missing Value Check
library(tidyverse) #loaded tidyverse

NHIS_Data_2021 <- read.csv("Data/NHIS_Data_2021.csv") #loaded dataset
summary(NHIS_Data_2021) #loaded summary of data to see range of variables

NHIS_vars <- NHIS_Data_2021 %>% #created subset of cleaned dataset with only needed variables
  select(AGEP_A, WEIGHTLBTC_A, HEIGHTTC_A, SEX_A, HISPALLP_A, EDUCP_A, PHSTAT_A, LSATIS4R_A) %>% #selected only variables needed
  filter( #used filter to remove missing values per codebook #! removes unwanted values
    AGEP_A >= 18 & AGEP_A <= 85 & !AGEP_A %in% c(97, 98, 99), #Age range from 18-85
    WEIGHTLBTC_A >= 100 & WEIGHTLBTC_A <= 299 & !WEIGHTLBTC_A %in% c(996, 997, 998, 999), #Weight range from 100-299
    HEIGHTTC_A >= 59 & HEIGHTTC_A <= 76 & !HEIGHTTC_A %in% c(96, 97, 98, 99), #Height range from 59-76
    SEX_A >= 1 & SEX_A <= 2 & !SEX_A %in% c(7, 9), #Sex range from 1-2
    HISPALLP_A >= 1 & HISPALLP_A <= 7 & !HISPALLP_A %in% c(97, 98, 99), #Race/ethnicity range from 1-7
    EDUCP_A >= 0 & EDUCP_A <= 10 & !EDUCP_A %in% c(97, 98, 99), #Education range from 0-10
    PHSTAT_A >= 1 & PHSTAT_A <= 5 & !PHSTAT_A %in% c(7, 8, 9), #General health status range from 1-5
    LSATIS4R_A >= 1 & LSATIS4R_A <= 4 & !LSATIS4R_A %in% c(7, 8, 9)) #Quality of life range from 1-4

summary(NHIS_vars) #loaded summary of cleaned data to verify all variables are within codebook range & missing values removed

#Task 2: Recoding Variables
NHIS_clean <- NHIS_vars %>% #created final cleaned dataset and added recoded education variable
  mutate(
    EDUCP_recode = case_when( #recoded education variable based on NHIS codebook categories
      EDUCP_A >= 0 & EDUCP_A <= 3 ~ 1, #0-3 Less Than High School
      EDUCP_A == 4 ~ 2, #4 High School Graduate
      EDUCP_A >= 5 & EDUCP_A <= 7 ~ 3, #5-7 Some College Education
      EDUCP_A >= 8 & EDUCP_A <= 10 ~ 4 #8-10 College Graduate or Better
    )
  )

NHIS_clean <- NHIS_clean %>%
  mutate(
    EDUCP_recode = factor( #converted recoded education variable into a labeled factor
      EDUCP_recode,
      levels = c(1, 2, 3, 4), #defined category order
      labels = c( #labeled each category in order
        "Less than High School",
        "High School Graduate",
        "Some College Education",
        "College Graduate or Better"
      )
    )
  )

summary(NHIS_clean) #loaded summary of cleaned dataset to confirm recoding worked.

write.csv(NHIS_clean, "Data/NHIS_clean.csv", row.names = FALSE) #saved final cleaned dataset as CSV file

```

# Day 3: Univariate & Bivariate Analysis & Visualization

```{r}

library(tidyverse) #load tidyverse packages
NHIS_clean <- read.csv("Data/NHIS_clean.csv")  #import dataset

summary(NHIS_clean$AGEP_A) #summary stats on mean,median, mode...
summary(NHIS_clean$WEIGHTLBTC_A) #summary stats for weight
summary(NHIS_clean$HEIGHTTC_A)  #summary stats for height

#Extra Stats
mean(NHIS_clean$AGEP_A, na.rm = TRUE)  #average age while removing missing values
sd(NHIS_clean$AGEP_A, na.rm = TRUE)  #standard deviation

#Histograms
hist(NHIS_clean$AGEP_A, main="Age Distribution", xlab="Age") #histogram for age
hist(NHIS_clean$WEIGHTLBTC_A, main="Weight Distribution", xlab="Weight")
hist(NHIS_clean$HEIGHTTC_A, main="Height Distribution", xlab="Height")
```
The code loads the dataset and explores key variables (age, weight, and height) by generating summary statistics, calculating additional measures for age, and creating histograms to visualize their distributions.The histogram shows that most weight values cluster between about 120 and 250 pounds, indicating a right-skewed distribution, while a small number of extremely high values near 1000 pounds suggest strong outliers or possible data entry errors.

```{r}
ggplot(NHIS_clean, aes(x = AGEP_A)) + #plot using dataset
  geom_histogram(bins = 30) + #histogram within 30 bins
  labs(title="Age Distribution", x="Age", y="Count")  #title 
```

The code uses ggplot2 to create a histogram of the age variable, displaying its distribution with labeled axes and a descriptive title.The histogram shows that ages are fairly spread out across a wide range (roughly 20 to 80), indicating a relatively even distribution, with a slight concentration in middle-aged groups and a few extreme values near the upper end suggesting potential outliers.

```{r}
ggplot(NHIS_clean, aes(x = WEIGHTLBTC_A)) + #ggplot
  geom_histogram(bins = 30) +  #histogram of weight within 30 bins
  labs(title="Weight Distribution", x="Weight", y="Count") #add tittle
```

The code uses ggplot2 to visualize the distribution of weight by creating a histogram with appropriate labels and a descriptive title.The weight distribution is slightly right-skewed, with most values concentrated between about 130 and 220 pounds. There are fewer higher-weight observations extending toward 300 pounds, indicating a gradual tail on the upper end.

```{R}
ggplot(NHIS_clean, aes(x = HEIGHTTC_A)) + #ggplot
  geom_histogram(bins = 30) + #histogram of height within 30 bins
  labs(title="Height Distribution", x="Height", y="Count") #add tittle
```

The code uses ggplot2 to create a histogram that visualizes the distribution of height, with clear axis labels and a descriptive title.The histogram shows that most height values are concentrated between about 60 and 72 inches, indicating a roughly normal distribution, while a small group of unusually high values (around 95–100 inches) suggests the presence of potential outliers or data entry errors.

```{r}
#Boxplots
boxplot(NHIS_clean$AGEP_A, main="Age Boxplot") #boxplot for age
boxplot(NHIS_clean$WEIGHTLBTC_A, main="Weight Boxplot") #boxplot for weight
boxplot(NHIS_clean$HEIGHTTC_A, main="Height Boxplot") #boxplot for height
```

The code generates boxplots for age, weight, and height to visualize their distributions, highlighting the median, spread, and any potential outliers in each variable.The median weight is around 170 pounds, with most values falling between roughly 150 and 200 pounds. There are several high-end outliers above about 280 pounds, indicating a slight right skew in the distribution.

```{r}
ggplot(NHIS_clean, aes(y = AGEP_A)) + #ggplot on y axis
  geom_boxplot() + #boxplot for age including median, quartiles, outliers
  labs(title="Age Boxplot") #add tittle
```

The code uses ggplot2 to create a boxplot of age, allowing visualization of its distribution, central tendency, and potential outliers.The median age is around the mid-50s, with the middle 50% of values roughly between about 40 and 70 years. The distribution appears fairly symmetric with no strong outliers, indicating ages are relatively evenly spread.

```{r}
ggplot(NHIS_clean, aes(y = WEIGHTLBTC_A)) + #ggplot on y axis
  geom_boxplot() + #box plot to distribute data
  labs(title="Weight Boxplot") #tittle
```

The code uses ggplot2 to create a boxplot of weight, visualizing its distribution, central tendency, and any potential outliers.The median weight is around 170–175 pounds, with the middle 50% of values roughly between about 150 and 200 pounds. There are several high-value outliers near 280–300 pounds, indicating a right-skewed distribution.

```{r}
ggplot(NHIS_clean, aes(y = HEIGHTTC_A)) + #ggplot to y axis
  geom_boxplot() + #distribution to height
  labs(title="Height Boxplot") #add tittle
```

The code uses ggplot2 to create a boxplot of height, allowing visualization of its distribution, central tendency, and potential outliers.The median height is around 66 inches, with the middle 50% of values roughly between 64 and 70 inches. The distribution appears fairly symmetric with no noticeable outliers, indicating consistent height values.

```{r}
#Qualitative Variables
#Frequency Tables
table(NHIS_clean$SEX_A)
table(NHIS_clean$HISPALLP_A)
table(NHIS_clean$EDUCP_recode)
table(NHIS_clean$PHSTAT_A)
table(NHIS_clean$LSATIS4R_A)
```
The frequency tables show that there are 26,037 participants in the sample dataset. From those participants, females have a slightly higher representation. The majority of respondents are Non-Hispanic White. The majority of individuals have an education College Graduate or Higher. Most participants reported Good to Very Good general health. The majority of participants reported being Very Satisfied with life.
```{r}
#Bar Plots
barplot(table(NHIS_clean$SEX_A), main="Sex Distribution")
barplot(table(NHIS_clean$HISPALLP_A), main="Race/Ethnicity")
barplot(table(NHIS_clean$EDUCP_recode), main="Education Level")
barplot(table(NHIS_clean$PHSTAT_A), main="Health Status")
barplot(table(NHIS_clean$LSATIS4R_A), main="Life Satisfaction")
```

```{r}
ggplot(NHIS_clean, aes(x = factor(SEX_A))) +
  geom_bar() +
  labs(title="Sex Distribution", x="Sex", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(HISPALLP_A))) +
  geom_bar() +
  labs(title="Race/Ethnicity", x="Category", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(EDUCP_recode))) +
  geom_bar() +
  labs(title="Education Level", x="Category", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(PHSTAT_A))) +
  geom_bar() +
  labs(title="Health Status", x="Category", y="Count")
```
```{r}
ggplot(NHIS_clean, aes(x = factor(LSATIS4R_A))) +
  geom_bar() +
  labs(title="Life Satisfaction", x="Category", y="Count")
```
The barplots show that for Sex Distribution, Females have a higher count. For Race/Ethnicity, Non-Hispanic White is over represented, with more than 3 times the count of the second highest category, Hispanic. For Education Level, the lowest category is Less Than High School, with College Graduate or Better almost 5 times more. The 2 main Health Statuses reported are Good and Very Good Health. The majority of participants reported being Satisfied and Very Satisfied with life, with very few reporting Dissatisfied and even less Very Dissatisfied.


```{r}
#04_bivariate_analysis.R 

library(tidyverse)
NHIS_clean <- read.csv("Data/NHIS_clean.csv")

#Quantitative vs. Qualitative:
boxplot(AGEP_A ~ SEX_A, #create boxplot for varbiables
        data = NHIS_clean,
        main = "Age by Sex", # title
        xlab = "Sex", # x axis label
        ylab = "Age") #y axis label
```
Key

Male: 1 

Female: 2

Age distributions are similar across both sex groups, with comparable medians and spread, suggesting no significant difference in age by sex.

```{r}
boxplot(AGEP_A ~ EDUCP_recode, #create boxplot for variables
        data = NHIS_clean,
        main = "Age by Education Level", #title
        xlab = "Education Level", # x axis label
        ylab = "Age") #y axis label
```
The boxplot shows age changes a little by education level, but they’re still pretty similar overall. Higher education groups look like they have slightly higher median ages, while lower education groups are a bit younger. There’s a lot of overlap though so the difference isn’t that big.

```{r}
ggplot(NHIS_clean, aes(x = factor(SEX_A), y = AGEP_A)) + # create ggplot for varibles
  geom_boxplot() + # assign boxplot
  labs(title = "Age by Sex", #title
       x = "Sex", # x axis variable
       y = "Age") # y axis variable
```
Key 

Male: 1
Female: 2

The boxplot shows that age between the two sex groups is pretty similar overall. Both groups have around the same median age, maybe slightly higher for one group but not by much.

```{r}
ggplot(NHIS_clean, aes(x = factor(EDUCP_recode), y = AGEP_A)) + #create ggplot for varaible
  geom_boxplot() + # assign boxplot to ggplot
  labs(title = "Age by Education Level", #create title
       x = "Education Level", # x axis 
       y = "Age") + coord_flip() # y axis
```
The boxplot shows that age is pretty similar across all education levels. There are some small differences, like slightly higher median ages in certain groups, but overall the distributions overlap a lot. This suggests education level doesn’t really change age that much in this dataset.

```{r}
#Qualitative vs. Qualitative:
ggplot(NHIS_clean, # create ggplot for variables 
       aes(x = factor(PHSTAT_A), fill = factor(LSATIS4R_A))) +
  geom_bar(position = "dodge") + # assign as a bar graph
  labs(title = "General Health vs Life Satisfaction", #title
       x = "General Health", #x axis 
       y = "Count", # y axis 
       fill = "Life Satisfaction") # Key on the side
```
Key

1: Excellent
2: Very good
3: Good
4: Fair
5: Poor

The chart shows a clear relationship between general health and life satisfaction. People who report excellent or very good health (1–2) tend to have higher life satisfaction, while those with fair or poor health (4–5) more often report lower satisfaction. There’s still some overlap, but overall better health is linked to better life satisfaction.

```{r}
#Quantitative vs. Quantitative:
plot(NHIS_clean$HEIGHTTC_A, # simple plot graph between two variables
     NHIS_clean$WEIGHTLBTC_A,
     main = "Height vs Weight", #title
     xlab = "Height", # x axis 
     ylab = "Weight") # y axis
```
The scatterplot shows a positive relationship between height and weight. As height increases, weight also tends to increase, although there is a lot of variation at each height. The points are pretty spread out, so it’s not a perfect relationship but the general trend is still clear.

```{r}
ggplot(NHIS_clean, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A)) + #ggplot of two variables
  geom_point(alpha = 0.5) + # assign as a scatter plot point graph
  labs(title = "Height vs Weight", #title 
       x = "Height", #x axis
       y = "Weight") # y axis
```
```{r}
cor(NHIS_clean$HEIGHTTC_A,
    NHIS_clean$WEIGHTLBTC_A,
    use = "complete.obs")
```

There is a moderate positive relationship between height and weight, indicating that as height increases, weight also tends to increase. The correlation coefficient (r = 0.50) suggests a moderate linear association between the two variables.


# Day 4: Multivariate Analysis & Report Drafting

```{r}
# Select variables and remove missing values
multi_data <- NHIS_clean %>%
  select(HEIGHTTC_A, WEIGHTLBTC_A, SEX_A, EDUCP_recode, PHSTAT_A, AGEP_A) %>%
  na.omit()

# Convert variables to factors 
multi_data$SEX_A <- as.factor(multi_data$SEX_A)
multi_data$EDUCP_recode <- as.factor(multi_data$EDUCP_recode)
multi_data$PHSTAT_A <- as.factor(multi_data$PHSTAT_A)

# Scatterolot of height vs weight, colored by sex and faceted by education 
ggplot(multi_data, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A, color = SEX_A)) +
  geom_point(alpha = 0.6) + # points with transparency 
  facet_wrap(~ EDUCP_recode) + #split by education level
  labs(
    title = "Height vs Weight by Sex and Education",
    x = "Height (cm)",
    y = "Weight (lbs)",
    color = "Sex"
  ) +
  theme_minimal() # clean theme

```
The scatterplots show a positive relationship between height and weight across all education levels. This pattern looks pretty consistent regardless of education, and both sex groups follow a similar trend. There are some small differences, but overall the relationship between height and weight stays about the same.



